---
title: 'Case 1: Bike Share'
author: "Fernando Cordero"
date: "10/16/2021"
output:
  html_document: default
  pdf_document: default
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{=html}
<style>
body {
text-align: justify}
</style>
```
# **Executive summary**

This report describes the analysis of 12 months of rides on bikes owned by Cyclistic, a bike share company based in Chicago, Il.
This analysis includes rides started between April 2019 and March 2020, and the objective is to determine potential ways to convert casual riders (customers on single-ride or day-pass plans) into subscribers (customer's on annual plans) by determining how these two user groups use the Cyclistic system differently.
The analysis finds that subscribers use the system much more frequently during the week, while casual riders complete many more rides during the weekend, and that subscriber's rides are far shorter in term of duration than casual rider's rides.
This analysis, however, should be treated as an exploratory analysis, and further analysis should be conducted in order to increase the certainty level of the findings discussed in this report, and more information about Cyclistic should be provided to make concrete recommendations.

# **Scenario and statement of purpose**

Cyclistic (or the Company) is a fictitious bike share business operating in Chicago since 2016.
Until now, Cyclistic's business strategy was to increase awareness of the Company and expand the business' clientele by having multiple payment plans and rental options.
Cyclistic's financial analysis team determined that customers that have an annual plan (subscribers) are more profitable than single-day-pass or single-ride-pass users (casual riders or customers - 'customer' is used interchangeably with casual rider it is a single word identifier, that facilitates data wrangling).
Cyclistic's leadership is interested in converting casual riders into subscribers.
This analysis will focus on: (1) determining key differences in the use pattern of the two user groups described above: subscribers and casual riders; (2) determining potential motives for casual riders to become subscribers; and (3) determine potential marketing strategies that can be used to influence casual riders into becoming subscribers.

## Business Task

The business task of this analysis is to determine potential ways to convert Cyclistic's clients from casual riders into annual members by exploring differences in the use patterns of these two user groups.

# **The Data**

The dataset used for this project is owned by the city of Chicago, IL., made public and available by Motivate International Inc. ("Motivate") under this [licence agreement](https://www.divvybikes.com/data-license-agreement).

## About the data

As Motivate is the operator of this bike share system and provider of the data, this dataset is considered to be first-hand.
This report will treat the data as if it was originated by the fictitious company Cyclistc.
The raw data provided for this analysis project contained the following information: Ride_id: containing an unique alphanumeric identifier Start_time and end_time: timestamps for pickup and drop off times Station_name and station_id: the pickup and drop off stations names and a numeric id that references a unique station name User_type: Boolean value, showing if the user of each ride is a casual rider (customer) or a subscriber The dataset used was stored in files that contained data organized in year-quarters.
It was combined, wrangled and cleaned up for analysis using R.
It is worth mentioning that while some of the datafiles used contained more variables, however, when these variables were not present in all datafiles, it was ignored.

## Cleaning process

After combining all data sets, the data was cleaned under the following criteria: Ride_id: Confirmed that all samples are unique, and that the characters are all alphanumeric with no unexpected characters.
Dates: Confirmed that start dates are all within the expected dates (between 4/1/2019 at 00:00:00 AM and 3/31/2020 at 11:59:59 PM) and that all start dates are prior to the corresponding end date; added two columns for ride duration (difference between start time and end time), one calculated in seconds and the other in decimal hours; finally, to make visualizations in R easier, columns were added for the year, month, day of the week, and part of the day (morning, afternoon, evening and night) for both start time and end time.
Stations: Confirmed that all station names unique values refer to only one station id unique value for both start and end stations; delete data points where both the station name and station id for either the start or end is missing, and remove station names that refer HQ, as it is explained in the scenario that these are not real rides, but test performed by Cyclistc's staff.
User type: Consolidate all tags names for the user type into 'customer' and 'subscriber only.

[This script](https://github.com/fccorder/googlecert_case1_bike-share/blob/master/case1_bike_share_clean_script.R) was used to clean up the data.

# **Analysis**

```{r setup environment, message=FALSE}
# Load packages
library(tidyverse)
library(lubridate)
library(skimr)
library(ggplot2)

# Initialize functions and options
'%!in%' <- Negate('%in%')
options(scipen = 999)
```

## Summary of Analysis

After cleaning the data, the analysis focused on comparing the use patterns of subscribers and casual riders on three main aspects: the number of rides, the duration of rides, and the time of day in which rides happen more often.
During analysis, charts and plots were plotted to identify trends more quickly, alongside descriptive tables showing the numbers represented in the charts for more precise information.

## Use patterns

This chapter presents the charts and figures that describe the use patterns of casual riders and subscribers identified in this analysis and a summary of the findings.

```{r read data, echo=FALSE, message=FALSE}
df <- read_csv('Divvy_Trips_2019Q2_to_2020Q1.csv')
df <- df %>% mutate(
  duration_asdate = difftime(end_time, start_time, units = 'secs') %>% 
    as.numeric() %>% as_datetime(), #For graphics
  start_hour = df$start_time %>% format('%H:%M:%S'),
  start_year = df$start_time %>% year() %>% as_factor(),
  start_month = df$start_time %>% month(label = TRUE),
  start_day = df$start_time %>% day(),
  start_wday = df$start_time %>% wday(label = TRUE),
  end_hour = df$end_time %>% format('%H:%M:%S'),
  end_year = df$end_time %>% year() %>% as_factor(),
  end_month = df$end_time %>% month(label = TRUE),
  end_day = df$end_time %>% day(),
  end_wday = df$end_time %>% wday(label = TRUE)
)
df <- df %>% mutate(
  start_pday = case_when(
    start_hour >= '05:00:00' & start_hour < '12:00:00' ~ 'Morning',
    start_hour >= '12:00:00' & start_hour < '18:00:00' ~ 'Afternoon',
    start_hour >= '18:00:00' & start_hour < '22:00:00' ~ 'Evening',
    start_hour >= '22:00:00' | start_hour < '05:00:00' ~ 'Night') %>% 
    factor(levels = c('Morning', 'Afternoon', 'Evening', 'Night')),
  end_pday = case_when(
    end_hour >= '05:00:00' & end_hour < '12:00:00' ~ 'Morning',
    end_hour >= '12:00:00' & end_hour < '18:00:00' ~ 'Afternoon',
    end_hour >= '18:00:00' & end_hour < '22:00:00' ~ 'Evening',
    end_hour >= '22:00:00' | end_hour < '05:00:00' ~ 'Night') %>% 
    factor(levels = c('Morning', 'Afternoon', 'Evening', 'Night'))
)
```

### Number of rides
Over the analysed period, casual riders registered 902,182 rides, while subscribers had 2,973,859. There were over 3 times more subscriber rides than casual rider rides over this period.
The below figure shows the difference in the number of rides per month. In each month of the year, subscribers do more rides than casual riders: twice as many rides in the summer and up to 18 times as many rides in the winter. It is also clear that there are many more rides completed between April and October, the warmer months, for both subscribers and casual riders (year-long patterns).

```{r rides per month}
df %>% group_by(
  start_month, user_type) %>% count() %>% 
  pivot_wider(names_from = user_type, values_from = n) %>% 
  mutate(ratio = Subscriber / Customer)

ggplot(data = df) +
  geom_bar(
    mapping = aes(x = start_month, fill = user_type),
    position = 'dodge') +
  ggtitle('Rides per month') +
  labs(x = 'Months', y = 'Number of rides') +
  scale_fill_discrete(name = 'User type') +
  theme_minimal() +
  theme(plot.title = element_text(size = 20,
                                  hjust = 0.5,
                                  face = 'bold'))
```

The next chart shows that subscribers have more daily rides during weekdays, while casual rides are showing more daily rides during the weekends (week-long pattern). In terms of proportions, subscribers take between 3 and 5 times as many rides as casual riders do during the week, and between 35% and 50% more during the weekends.

```{r per day of the week}
df %>% group_by(start_wday, user_type = user_type) %>% count() %>% 
  pivot_wider(names_from = user_type, values_from = n) %>%
  mutate(ratio = Subscriber / Customer)

ggplot(data = df) +
  geom_bar(
    mapping = aes(x = start_wday, fill = user_type),
    position = 'dodge') +
  ggtitle('Rides per day of the week') +
  labs(x = 'Days', y = 'Number of rides') +
  scale_fill_discrete(name = 'User type') +
  theme_minimal() +
  theme(plot.title = element_text(size = 20,
                                  hjust = 0.5,
                                  face = 'bold'))
```

The following chart shows how the week-long pattern is repeated every month of the year. At the same time, the week-long tendencies each month are affected by the year-long tendency: the difference in the number of daily rides between weekdays and weekend days far greater during the summer months than during the winter, for both subscribers and casual riders.

```{r per month per day}
ggplot(data = df) +
  geom_bar(
    mapping = aes(x = start_wday, fill = user_type),
    position = 'dodge') +
  facet_wrap(~ start_month) +
  ggtitle('Rides per month per day') +
  labs(y = 'Number of rides') +
  scale_fill_discrete(name = 'User type') +
  theme_minimal() +
  theme(plot.title = element_text(size = 20,
                                  hjust = 0.5,
                                  face = 'bold'),
        axis.text.x = element_text(angle = 90),
        axis.title.x = element_blank(),
        legend.position = c(0,1.115))
```

When comparing each weekday throughout the year, it can be seen that every day of the week reflects the year-long tendency. There are two observations worth noting: (1) the difference in daily subscriber rides and customer rides is much more significant during weekdays, while during weekends the number of rides is very similar for both types of customers; and (2) casual rides registered more rides than subscribers only over all the accumulated Saturdays of July and August.

```{r rides per day per month}
ggplot(data = df) +
  geom_bar(
    mapping = aes(x = start_month, fill = user_type),
    position = 'dodge') +
  facet_wrap(~ start_wday) +
  ggtitle('Rides per day per month') +
  labs(y = 'Number of rides') +
  scale_fill_discrete(name = 'User type') +
  theme_minimal() +
  theme(plot.title = element_text(size = 20,
                                  hjust = 0.5,
                                  face = 'bold'),
        axis.text.x = element_text(angle = 90),
        axis.title.x = element_blank(),
        legend.position = c(0.9,0.07))
```

### Durations
Duration distribution through the year show that 95% of subscribers' rides are shorter than 40 minutes during the summer months and less than 30 minutes during the winter months; their duration does not vary much through the months. Casual rider’s ride duration, on the other hand, varies much more, with over 50% of casual rider’s rides being longer than subscriber rides, and rides are considerably longer during warmer months.

```{r duration distribution per month}
aggregate(
  df$duration_asdate ~ df$user_type + df$start_month, 
  FUN = mean) %>% 
  setNames(c('user_type', 'month', 'duration_asdate')) %>% 
  mutate(duration_asdate = duration_asdate %>% format('%m-%d %H:%M:%S')) %>% 
  pivot_wider(names_from = user_type, values_from = duration_asdate)

ggplot(data = df) +
  geom_boxplot(mapping = aes(x = start_month,
                             y = duration_asdate,
                             fill = user_type),
               outlier.shape = NA) +
  ggtitle('Ride duration per month') +
  labs(x = 'Month', y = 'duration in hours') +
  scale_fill_discrete(name = 'User type') +
  theme_minimal() +
  theme(plot.title = element_text(size = 20,
                                  hjust = 0.5,
                                  face = 'bold'))
```

Subscribers' ride duration is very similar between weekdays and weekend days, while casual riders' rides are longest on weekend days and get shorter towards the middle of the week.

```{r duration per day of the week}
aggregate(
  df$duration_asdate ~ df$user_type + df$start_wday, 
  FUN = mean) %>% 
  setNames(c('user_type', 'day', 'duration_asdate')) %>% 
  mutate(duration_asdate = duration_asdate %>% format('%m-%d %H:%M:%S')) %>% 
  pivot_wider(names_from = user_type, values_from = duration_asdate)

ggplot(data = df) +
  geom_boxplot(mapping = aes(x = start_wday,
                             y = duration_asdate,
                             fill = user_type),
               outlier.shape = NA) +
  ggtitle('Ride duration per day') +
  labs(x = 'Month', y = 'Duration in hours') +
  scale_fill_discrete(name = 'User type') +
  theme_minimal() +
  theme(plot.title = element_text(size = 20,
                                  hjust = 0.5,
                                  face = 'bold'))
```

The following figure shows that ride durations follow a week-long pattern that can be seen every month for both subscribers and casual riders. This tendency is affected by the year-long tendency: rides are far longer during the summer months than winter months, while subscriber’s ride durations are very close in the summer and winter months.

```{r duration per month per day}
ggplot(data = df) +
  geom_boxplot(mapping = aes(x = start_wday,
                             y = duration_asdate,
                             fill = user_type),
               outlier.shape = NA) +
  facet_wrap(~ start_month) +
  ggtitle('Ride duration per month per day') +
  labs(y = 'Duration in hours') +
  scale_fill_discrete(name = 'User type') +
  theme_minimal() +
  theme(plot.title = element_text(size = 20,
                                  hjust = 0.5,
                                  face = 'bold'),
        axis.text.x = element_text(angle = 90),
        axis.title.x = element_blank(),
        legend.position = c(0,1.115))
```

```{r duration per day per month}
ggplot(data = df) +
  geom_boxplot(mapping = aes(x = start_month,
                             y = duration_asdate,
                             fill = user_type),
               outlier.shape = NA) +
  facet_wrap(~ start_wday) +
  ggtitle('Ride duration per day per month') +
  labs(y = 'Duration in hours') +
  scale_fill_discrete(name = 'User type') +
  theme_minimal() +
  theme(plot.title = element_text(size = 20,
                                  hjust = 0.5,
                                  face = 'bold'),
        axis.text.x = element_text(angle = 90),
        axis.title.x = element_blank(),
        legend.position = c(0.9,0.07))
```

For each day duration is noticeably affected by the year-long tendency: rides are longer during summer and warmer months for both user groups. The exemption to this is Wednesday in March, but when looking closer it seems to be an exception to the rule on .

```{r excepction wedneday in march}
ggplot(data = df %>% filter(start_wday == 'Wed' & start_month == 'Mar')) +
  geom_boxplot(mapping = aes(x = start_day %>% as.factor(),
                             y = duration_asdate,
                             fill = user_type),
               outlier.shape = NA) +
  ggtitle('Ride duration per day, Wednesdays in March') +
  labs(x = 'date in September', y = 'Duration in hours') +
  scale_fill_discrete(name = 'User type') +
  theme_minimal() +
  theme(plot.title = element_text(size = 20,
                                  hjust = 0.5,
                                  face = 'bold'))
```

### Pick up and drop offs
There are many more subscriber rides during the morning and afternoon during weekend days, coinciding with commuting times, while there is a spike in casual rider’s rides on weekend afternoons.

```{r pick ups and drop off per day of the week}
# Starts
ggplot(data = df) +
  geom_bar(
    mapping = aes(x = start_pday, fill = user_type),
    position = 'dodge') +
  facet_wrap(~ start_wday) +
  ggtitle('Pick up times per day') +
  labs(y = 'Count') +
  scale_fill_discrete(name = 'User type') +
  theme_minimal() +
  theme(plot.title = element_text(size = 20,
                                  hjust = 0.5,
                                  face = 'bold'),
        axis.text.x = element_text(angle = 90),
        axis.title.x = element_blank())
# Ends
# Per day of the week:
ggplot(data = df) +
  geom_bar(
    mapping = aes(x = end_pday, fill = user_type),
    position = 'dodge') +
  facet_wrap(~ end_wday) +
  ggtitle('Drop off times per day') +
  labs(y = 'Count') +
  scale_fill_discrete(name = 'User type') +
  theme_minimal() +
  theme(plot.title = element_text(size = 20,
                                  hjust = 0.5,
                                  face = 'bold'),
        axis.text.x = element_text(angle = 90),
        axis.title.x = element_blank()) 
```

The next charts show there are very few rides during summer nights for both user groups, decrease significantly on colders months and are extremely rare during winter nights. There seems to be a core group of subscribers that continues to ride through the winter months on mornings and afternoons, but between April and October there is a very evident increment in the number of users following seasonal trends. For casual riders, significantly more rides start during the afternoon between April and October, and between November and February there are very few trips at all.

```{r pick ups and drop off per month}
# Starts
ggplot(data = df) +
  geom_bar(
    mapping = aes(x = start_pday, fill = user_type),
    position = 'dodge') +
  facet_wrap(~ start_month) +
  ggtitle('Pick up times per month') +
  labs(y = 'Count') +
  scale_fill_discrete(name = 'User type') +
  theme_minimal() +
  theme(plot.title = element_text(size = 20,
                                  hjust = 0.5,
                                  face = 'bold'),
        axis.text.x = element_text(angle = 90),
        axis.title.x = element_blank()) 

# Ends
ggplot(data = df) +
  geom_bar(
    mapping = aes(x = end_pday, fill = user_type),
    position = 'dodge') +
  facet_wrap(~ end_month) +
  ggtitle('Drop off times per month') +
  labs(y = 'Count') +
  scale_fill_discrete(name = 'User type') +
  theme_minimal() +
  theme(plot.title = element_text(size = 20,
                                  hjust = 0.5,
                                  face = 'bold'),
        axis.text.x = element_text(angle = 90),
        axis.title.x = element_blank())
```

For further analysis and customizable detail, see [these interactive charts](https://public.tableau.com/views/Bikeshare_case1/Dashboard1?:language=en-US&:display_count=n&:origin=viz_share_link) comparing the use patterns of casual riders and subscribers.

# **Conclusion**

Based on the analysis described in this report, only general recommendations can be made as essential background information was not provided in the scenario statement nor can this information be assumed. Even though the data used for the analysis is captured in the real world and based on a company similar to Cyclistic, the missing information is:

- Pricing plans: Although it is known that Cyclist offers multiple pricing plans, there is no clarification on pricing, nor on the rules and restrictions for each plan.
- Type of bike: The scenario statement mentions that there are 3 types of bike that Cyclist makes available to their users. This information was only included for rides that occurred during the last 3 months of data included in this analysis.
 - Only one year (2019 Q2 to 2020 Q1) was analysed. Further analysis should be done over a longer period of time. An analysis performed over a longer period of time would increase certainty on the results of the analysis . Moreover, the Covid-19 pandemic started in 2019-Q1, which could have affected the use patterns of both casual riders and subscribers.
 
## Key findings and recommendations

- At least 50% of casual rider’s rides are longer than the majority of subscriber’s rides.
- Casual rider rides are always longer, and more numerous during weekend days than during weekdays, with most of those rides occuring during the afternoon. Subscriber rides, on the other hand, keep similar duration every day of the week, and there are more rides during weekdays.
- Both user type rides are affected by the seasons through the year: However, while casual rider rides can get to fewer than 10,000 monthly rides, subscriber monthly rides don’t get below 115,000 during the winter. At the same time, casual rider rides are considerably shorter during the winter, while subscriber rides duration are far less affected by the cold months.

Based on the key findings of this report, the recommendations are:

- If it does not already exist, create a subscription plan in which users can do longer rides, or that have convenient rates for weekends. This recommendation is vague due to the lack of context and information on the pricing plans offered by Cyclistic.
- To better inform the recommendations, more context should be provided:
  - Pricing plans: With this information, it would be possible to correlate the rules and conditions of the different plans with the use patterns of subscriber and casual riders.
  - Type of bike used: As mentioned before, Cyclistic offers 3 types of bikes to its customers but this information was not complete in the analysed dataset. It is important to determine differences in how subscribers and casual rides use all bike types.
  - User id: Adding information to determine how often user’s ride with Cyclistic
  - Geographical information: This would allow mapping trips. This information would help identify the purpose of use of both user groups.

- Increase confidence level of analysis:
  - As it was mentioned previously, this analysis was performed only on rides between April 2019 and March 2020. Furthermore, the Covid-19 pandemic started during the first quarter of 2020, which could have impacted the way users (both subscribers and casual riders) ride Cyclistic bikes.
  - During this analysis, several trends were identified and explained. However, this analysis should be repeated for a longer period to confirm or refute these trends. Moreover, analysing a longer period of time could enable predictions of Cyclistic’s growth, stress periods, etc.
  
This analysis should be treated as a first analysis/exploratory and should be supported with ensuing analysis with a larger scope. Further analysis should cover a longer period of time and include, as discussed above, more information to help increase the confidence level of the analysis. Finally,  more information about the context of the Company should be provided to help create a more clear picture of the use pattern of Cyclistic’s customers.
